\thetitle

### 2020-11-21 writing/discussion.tex
Fix missing section label, number of scenarios
typos + small edits

%% -----------------------------------------------------------------------------
\section{What We Learned}

@@

The analysis in the previous section suggests a number of interesting
high-level conclusions about blame in the gradual typing setting.  First,
run time type checks have a large positive impact. This is the case
regardless of whether these checks issue blame or throw plain exceptions.
Second, in many cases type checks that issue blame are more helpful than
those that do not. However, our results indicate that in other cases blame
is not critical and thus it is worth further investigation whether
disabling blame tracking for the sake of performance may be   an
acceptable trade off. Third, from the three different blame strategies, the
Natural approach of Typed Racket fares better than the Transient approach
of Reticulated Python, but only by a small margin. Since Natural offers complete
and sound path-based blame while Transient offers incomplete but sound heap-based
blame~\cite{gfd-oopsla-2019}, our results call for a deeper understanding
of the relative strengths of the two models for blame.  Fourth, given that
Transient's sound but shallow run time type checks do not seem to hamper
debugging, a version of Typed Racket that disables some of its wrappers
may offer an answer to the performance issues of the Natural semantics. 
Fifth, just as for Typed Racket, the experiment points to possible
improvements for Reticulated Python. In particular, the experimental 
data indicates that the usefulness of transient blame does not depend on
whether  the rational programmer follows first or the last element from a
Transient blame set. Hence, given the performance issues of Transient we discuss further on, 
a possible optimization is to limit the size of blame sets
using timestamps.

@@

As we discuss in section~\ref{sec:results}, Transient surprisingly blames
an empty set of boundaries in 967 scenarios.
Such cases should never occur, at least in theory, because an empty set
 means the value has never crossed a boundary---if the value is indeed defined
 in the current module, then we have typed code blaming itself for a typed
 value as all the checks are inlined in typed code. Thus, the type checker must be unsound.
After careful investigation of these empty blame cases, we found no soundness
 bugs in Transient Typed Racket.
Instead, we found scenarios in which Transient lost track of the proper
 boundaries.
These scenarios suggest ways to improve~\citet{vss-popl-2017}'s blame
algorithm:
\begin{itemize}
  \item
    Entries in a blame map must point to several parent entries.
    For example, if the function \texttt{f} receives bad input in the call
    \texttt{(filter f xs)}, then blame should point to both \texttt{filter}
    and the \texttt{xs} list.
  \item
    The construction of blame map entries must be guided by type-like specifications
     instead of relying on syntax.
    Aliasing the built-in \texttt{filter} function should not change the shape
     of the blame map.
  \item
    The initial blame map must reflect the initial type environment.
    For instance, Typed Racket trusts that untyped core-library functions, such as \texttt{filter},
     behave correctly; such assumptions should appear in the blame map.
\end{itemize}
\noindent{}Despite these known limitations, based on the relatively small
number of debugging failures, we conjecture that these improvements
 to Transient blame-tracking do not affect our overall conclusions.

A second apparent flaw in our Transient semantics is the high cost of blame.
Whereas \citet{vss-popl-2017} report an average slowdown of 2.5x and
worst-case of 5.4x on fully-typed benchmarks in Reticulated,
some of our configurations exceed our 4 minute timeout or 6GB memory limit with transient blame.
To put those limits into context, the fully typed and fully untyped benchmarks all complete in a few seconds with minimal memory usage, and none of the mixed Natural configurations hit these limits.
Unfortunately, our high cost for Transient seems closer to the truth;
 there are at least three broad issues that skew the earlier results.
First, the 2017 implementation of Reticulated fails to insert certain
 soundness checks\footnote{Missing check: \url{https://github.com/mvitousek/reticulated/issues/36}}
 and blame-map updates\footnote{Missing cast: \url{https://github.com/mvitousek/reticulated/issues/43}}
 from the paper.
Second, Reticulated's type checker assigns the dynamic type to local
 variables in the "fully typed" benchmarks---often because the type system
 cannot articulate a more precise type.
Code that ends up with the dynamic type has fewer constraints to check at run-time.
Third, \citet{vss-popl-2017} use relatively small benchmarks.
Four have since been retired from the official Python benchmark suite
 because they are too small, unrealistic, and unstable.\footnote{Release notes: \url{https://pyperformance.readthedocs.io/changelog.html}}
On larger programs, Reticulated suffers from high overhead.
For example, a Reticulated version of our simplest benchmark runs in
 ~40 seconds without blame, and with blame times out after 10 minutes.

>>>>

%% -----------------------------------------------------------------------------
\section{What We Learned}
\label{sec:discussion}

@@

The analysis in the previous section suggests a number of interesting
high-level conclusions about blame in the gradual typing setting.  First,
run time type checks have a large positive impact. This is the case
regardless of whether these checks issue blame or throw plain exceptions.
Second, in many cases type checks that issue blame are more helpful than
those that do not. However, our results indicate that in other cases blame
is not critical and thus it is worth further investigation whether
disabling blame tracking for the sake of performance may be   an
acceptable trade off. Third, from the three different blame strategies, the
Natural approach of Typed Racket fares better than the Transient approach
of Reticulated Python, but only by a small margin. Since Natural offers complete
and sound path-based blame while Transient offers incomplete but sound heap-based
blame~\cite{gfd-oopsla-2019}, our results call for a deeper understanding
of the relative strengths of the two models for blame.  Fourth, given that
Transient's sound but shallow run time type checks do not seem to hamper
debugging, a version of Typed Racket that disables some of its wrappers
may offer an answer to the performance issues of Natural. 
Fifth, just as for Typed Racket, the experiment points to possible
improvements for Reticulated Python. In particular, the experimental 
data indicates that the usefulness of transient blame does not depend on
whether  the rational programmer follows first or the last element from a
blame set. Hence, given the performance issues of Transient we discuss further on, 
a possible optimization is to limit the size of blame sets
using timestamps.

@@

As we discuss in section~\ref{sec:results}, Transient surprisingly
produces
an empty blame set in 967 scenarios.
Such cases should never occur, at least in theory, because an empty set
 means the value has never crossed a boundary---if the value is indeed defined
 in the current module, then we have typed code blaming itself for a typed
 value as all the checks are inlined in typed code. Thus, the type checker must be unsound.
After careful investigation of these empty blame cases, we found no soundness
 bugs in Transient Typed Racket.
Instead, we found scenarios in which Transient lost track of the proper
 boundaries.
These scenarios suggest ways to improve~\citet{vss-popl-2017}'s blame
algorithm:
\begin{itemize}
  \item
    Entries in a blame map must point to several parent entries.
    For example, if the function \texttt{f} receives bad input in the call
    \texttt{(filter f xs)}, then blame should point to both \texttt{filter}
    and the \texttt{xs} list.
  \item
    The construction of blame map entries must be guided by type-like specifications
     instead of relying on syntax.
    Aliasing the built-in \texttt{filter} function should not change the shape
     of the blame map.
  \item
    The initial blame map must reflect the initial type environment.
    For instance, Typed Racket trusts that untyped core-library functions, such as \texttt{filter},
     behave correctly; such assumptions should appear in the blame map.
\end{itemize}
\noindent{}Despite these known limitations, based on the relatively small
number of debugging failures, we conjecture that these improvements
 to Transient blame-tracking do not affect our overall conclusions.

A second apparent flaw in our Transient implementation is the high cost of blame.
Whereas \citet{vss-popl-2017} report an average slowdown of 2.5x and
worst-case of 5.4x on fully-typed benchmarks in Reticulated,
some of our configurations exceed our 4 minute timeout or 6GB memory limit with transient blame.
To put those limits into context, the fully typed and fully untyped benchmarks all complete in a few seconds with minimal memory usage, and none of the mixed Natural configurations hit these limits.
Unfortunately, our high cost for Transient seems closer to the truth;
 there are at least three broad issues that skew the earlier results.
First, the 2017 implementation of Reticulated fails to insert certain
 soundness checks\footnote{Missing check: \url{https://github.com/mvitousek/reticulated/issues/36}}
 and blame-map updates\footnote{Missing cast: \url{https://github.com/mvitousek/reticulated/issues/43}}
 from the paper.
Second, Reticulated's type checker assigns the dynamic type to local
 variables in the "fully typed" benchmarks---often because the type system
 cannot articulate a more precise type.
Code that ends up with the dynamic type has fewer constraints to check at run-time.
Third, \citet{vss-popl-2017} use relatively small benchmarks.
Four have since been retired from the official Python benchmark suite
 because they are too small, unrealistic, and unstable.\footnote{Release notes: \url{https://pyperformance.readthedocs.io/changelog.html}}
On larger programs, Reticulated suffers from high overhead.
For example, a Reticulated version of our simplest benchmark runs in
 ~40 seconds without blame, and with blame times out after 10 minutes.

### 2020-11-21 writing/landscape.tex
typos + small edits

The \texttt{untyped/pack-lib} module (left) provides, among
others, a function \texttt{pack} that consumes JSON data and packages it
in a dictionary. However, by mistake \texttt{pack} returns
a hash table instead of the association list its documentation promises.
The \texttt{client} module (right) uses \texttt{pack} indirectly to
package public and sensitive data it reads from the filesystem. It does so
through two different intermediary typed modules (center). The first,
\texttt{typed/pack-lib}, simply imports \texttt{pack} and immediately
exports as \texttt{typed-pack} asserting that it is a function
that consumes JSON and returns an association list that maps symbols to
strings. The second typed module, \texttt{typed/crypto-pack-lb},
also imports \texttt{pack} and assumes for it the same type as
\texttt{typed/pack-lb} but instead of exporting it directly it wraps in
another function, \texttt{crypto-pack}, that encrypts its input before
passing it to \texttt{pack}.

Under Natural semantics~\cite{tf-popl-2008,tfffgksst-snapl-2017}, when
\texttt{pack} is imported in either \texttt{typed/pack-lb} or
\texttt{typed/secure-pack-lb}, the runtime checks that \texttt{pack} is a
function and wraps it in a proxy that enforces its  type.  Analogously
the run time wraps the exports of the two typed modules in proxies that
check that untyped code applies \texttt{typed-pack} and
\texttt{crypto-pack} to arguments of its type. In essence, the
return type checks protect  the typed modules in case they use
\texttt{pack} and, as is the case here, it does not return an association
list. The check of the argument type  treats the
exports of the two typed modules as typed values that need to be protected
when they are used by untyped code.  As this analysis implies, if the
return type check fails then the problem is with
\texttt{untyped/pack-lib}, while if the argument type check fails
responsibility lies with \texttt{client}. The Natural semantics keep track
of the responsible party for each check in the corresponding proxy. Thus,
in our example, when the call to \texttt{typed-pack} returns, the return
type check fails and the Natural semantics blames \texttt{pack-lib}.

@@

Under Erasure semantics, the example does not raise an error due to the
gradual type system; Erasure performs no type checks at runtime. Thus the
program may fail with an exception if \texttt{client} tries to use the
result of \texttt{typed-pack} or \texttt{crypto-pack}, but it may also not
fail at all and produce an wrong result that goes undetected by the
gradual type system .  In either case, the Erasure semantics provides no
help to the programmer for detecting the bug in our program.

>>>>

The \texttt{untyped/pack-lib} module (left) provides, among
others, a function \texttt{pack} that consumes JSON data and packages it
in a dictionary. However, by mistake \texttt{pack} returns
a hash table instead of the association list its documentation promises.
The \texttt{client} module (right) uses \texttt{pack} indirectly to
package public and sensitive data it reads from the filesystem. It does so
through two different intermediary typed modules (center). The first,
\texttt{typed/pack-lib}, simply imports \texttt{pack} and immediately
exports as \texttt{typed-pack} asserting that it is a function
that consumes JSON and returns an association list that maps symbols to
strings. The second typed module, \texttt{typed/crypto-pack-lb},
also imports \texttt{pack} and assumes for it the same type as
\texttt{typed/pack-lb} but instead of exporting it directly \texttt{typed/pack-lb} wraps it in
another function, \texttt{crypto-pack}, that encrypts its input before
passing it to \texttt{pack}.

Under Natural semantics~\cite{tf-popl-2008,tfffgksst-snapl-2017}, when
\texttt{pack} is imported in either \texttt{typed/pack-lb} or
\texttt{typed/secure-pack-lb}, the runtime checks that \texttt{pack} is a
function and wraps it in a proxy that enforces its  type.  Analogously
the runtime wraps the exports of the two typed modules in proxies that
check that untyped code applies \texttt{typed-pack} and
\texttt{crypto-pack} to arguments of the expected type. In essence, the
return type checks protect  the typed modules in case they use
\texttt{pack} and, as is the case here, it does not return an association
list. The check of the argument type  treats the
exports of the two typed modules as typed values that need to be protected
when they are used by untyped code.  As this analysis implies, if the
return type check fails then the problem is with
\texttt{untyped/pack-lib}, while if the argument type check fails
responsibility lies with \texttt{client}. The Natural semantics keep track
of the responsible party for each check in the corresponding proxy. Thus,
in our example, when the call to \texttt{typed-pack} returns, the return
type check fails and the Natural semantics blames \texttt{pack-lib}.

@@

Under Erasure semantics, the example does not raise an error due to the
gradual type system; Erasure performs no type checks at runtime. Thus the
program may fail with an exception if \texttt{client} tries to use the
result of \texttt{typed-pack} or \texttt{crypto-pack}, but it may also not
fail at all and produce a wrong result that goes undetected by the
gradual type system .  In either case, the Erasure semantics provides no
help to the programmer for detecting the bug in our program.

### 2020-11-21 writing/rational.tex
Fix another typo
typos + small edits

When the buggy untyped component of a program is replaced by the typed counterpart,
the type checker fails if this component causes the impedance mismatch. Hence, a
trail that ends at an ill-typed scenarios successfully pinpoints the location of
bugs. 
\begin{quote}
\it A Natural blame trail $\conf_0,...\conf_n$ in a lattice $\lattice{\system}$ is
\emph{successful} iff its last scenario $\conf_n$ does not type check.  A Natural
blame trail $\conf_0,..,\conf_n$ in a lattice $\lattice{\system}$ is \emph{failing}
iff $\conf_n$ type checks and the trail cannot be extended further.
\end{quote}
That is, failing Natural blame trails are those that have not reached a scenario that
reveals the bug at compile time and their last scenario does not blame an untyped component. Thus
the rational programmer receives no hints from the gradual type system on how to
continue the search for the bug.

@@

The transient semantics assigns blame to a set of components. In this setting, a
blame assignment has the following meaning: the witness value to the impedance
mismatch may have crossed the boundaries between elements in the set, and each crossing
checked only a portion of the value's type shape.

@@

Furthermore, our experiment can compare the relative usefulness of blame
between any two of the three semantics. 
\begin{itemize}
\item[$Q_*$] Is blame for X more useful than blame for Y? (Where X and Y are any of Natural,
  Erasure or Natural?)
\end{itemize}

>>>>

When the buggy untyped component of a program is replaced by the typed counterpart,
the type checker fails if this component causes the impedance mismatch. Hence, a
trail that ends at an ill-typed scenarios successfully pinpoints the location of
bugs. 
\begin{quote}
\it A Natural blame trail $\conf_0,...\conf_n$ in a lattice $\lattice{\system}$ is
\emph{successful} iff its last scenario $\conf_n$ does not type check.  A Natural
blame trail $\conf_0,..,\conf_n$ in a lattice $\lattice{\system}$ is \emph{failing}
iff $\conf_n$ type checks and the trail cannot be extended further.
\end{quote}
That is, failing Natural blame trails are those that have not reached a scenario that
reveals the bug at compile time and their last scenario does not blame an untyped component. Thus
the rational programmer gets no hints from the gradual type system on how to
continue the search for the bug.

@@

The transient semantics assigns blame to a set of components. In this setting, a
blame assignment has the following meaning: the witness value to the impedance
mismatch may have crossed the boundaries between elements in the set, and each crossing
checked the value's type in a shallow manner.

@@

Furthermore, our experiment can compare the relative usefulness of blame
between any two of the three semantics. 
\begin{itemize}
\item[$Q_*$] Is blame for X more useful than blame for Y? (Where X and Y are any of Natural, Transient, or Erasure)
\end{itemize}
